Species-specific effects of the Urban Heat Island on tree growth across Berlin

### 2020-03-25 analysis/paper/paper.Rmd
cont work

---
title: "Species-specific effects of the Urban Heat Island on tree growth across Berlin"
author:
  - Alexander G. Hurley$^{a *}$, Ingo Heinrich$^a$
  - $^a$ *Geoforschungszentrum Potsdam, Section 4.3, Germany*
  - "* Corresponding author: hurley@gfz-potsdam.de"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::html_document2:
      fig_caption: yes
      toc: true
    bookdown::pdf_document2:
      fig_caption: yes
      toc: yes
bibliography: uhi_references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  **This document outlines the rationale for an analysis of tree growth (potential) and its relationship with the Urban Heat Island (UHI) effect in Berlin using an extensive, publicly available data set. It introduces preliminary results and provides an outlook for up-coming and potential work.**
keywords: |
  Uferbaeume; Anlagenbaeume; Strassenbaeume; 
highlights: |
  We will see. 
---

@@

To complement (existing) detailed dendroecological analyses of climate-growth relationships in Berlin for key species, we propose inferring growth modulation from a large data set in excess of 650000 individuals provided by the Berlin Senate Administration (Senatsverwaltung).
This data set contains information on location, species, trunk diameter (at breast height; $DBH$), and height, amongst other variables for street and park trees.
In a space-for-time substitution, growth of individual species can be assessed across the entire cite of Berlin, and related to effects of the UHI, while accounting for other location-specific factors, such as street characteristics, development intensity, available soil volume, etc.
Comparable applications are found, for example, in @quigley2004 and @pretzsch2017.
The former  inferred absolute growth potential for species across successional groups, and between rural and urban species, yet lacked spatially-explicit effect size estimates or predictions of maximum potential;
@pretzsch2017 applied linear hierarchical models to infer growth modulation for different cities, time periods and locations (urban $vs.$ rural).

**By contrast, we propose applying a statistical model that is spatially explicit, while also allowing to account for the nested nature of the data set (e.g. streets and districts) as well as other pertinent factors (hierarchical, generalized additive model, see Section$~$\@ref(sec:methods)).
This also allows to infer the absolute growth potential of species given, for example, a specific location, age or UHI magnitude.**
Tree-level growth data, however, is paramount in validating such relationships, and its inclusion in the model would also allow incorporating effects of varying climate.

@@

Currently, $DBH$ has been modelled using a hierarchical linear model (linear mixed effects model) with `lme4` [@bates2015] in @rcoreteam2020.
The general form of this model is:

\begin{equation}
Y_{i,j} = (\beta_0 + b_{0,i,j}) + (\beta_1 + b_{1,i,j}) \cdot x_i  + \epsilon_{i,j},
(\#eq:lme)
\end{equation}
where $\beta_0$ is the intercept with its random component $b_0$, and $\beta_1$ the slope with its random component $b_1$.
The random errors are assumed i.i.d. and distributed as $b \sim \mathcal{N}(0, \tau ^2)$. 
The model for which results are presented in Figure$~$\@ref(fig:fig-lmestat) estimates $DBH$ from tree age and the local UHI intensity as continuous covariates with random slopes and intercepts for each species;
note, that for computational efficiency each genera was modelled seperately.
Further, models were only established for species with at least 150 individuals.

@@

Additional remote sensing data for local context, higher resolution UHI data, and most important, temporally-resolved tree growth data are still required and/or would greatly improve the impact and confidence in conclusions.
See Tab.$~$\@ref(tab:tab-data-req) for details.

@@

@@

             caption = "Data requirements for analysis, including currently available/accessed and required/desired data.",
             caption.short = "Data requirements for analysis",
             col.names = c("Subject / Relevance",
                          "Desc.",
                          "Type",
                          "Obsv. (n)",
                          "Resolution (m)",
                          "Source")) %>% 
  kableExtra::kable_styling(full_width = TRUE, 
                            font_size = 11) %>% 
  kableExtra::row_spec(0, bold = TRUE, font_size = 12) %>% 
  kableExtra::row_spec(9,italic =  TRUE) %>% 
  kableExtra::pack_rows("available or accessed", 1, 4) %>%
  kableExtra::pack_rows("required and/or desired", 5, 9) %>% kableExtra::landscape()

@@

# Preliminary results

Tree locations are clustered and follow spatial structures based on their category, i.e. riparian, street and park trees (Fig.$~$\@ref(fig:fig-tree-overview-map)).
Planting in space and time shows species specific patterns (by districts), often related to major events, such as the start and end of wars, as well as the German separation (not shown here).
Table$~$\@ref(tab:tab-age) shows the binned distribution of generae across age classes.

@@

age_tables <- age_tables %>% 
  dplyr::arrange(desc(n_total))
colnames(age_tables)[c(1,7,8)] <- c("Genera", "Total (n)", "Missing (n)")

@@

if (grepl("docx", out_format) ||
    knitr::is_latex_output() ) {
  table_dat_req_kable <- age_tables %>% 
  knitr::kable(booktabs = TRUE,

             caption = "Trees",
             caption.short = "Trees"
             # col.names = c("Subject/Relevance",
             #              "Desc.",
             #              "Type",
             #              "Obsv. (n)",
             #              "Resolution (m)",
             #              "Source")
             ) %>% 
  kableExtra::row_spec(13, hline_after = TRUE)

} else {
table_dat_req_kable <- age_tables %>% 
  knitr::kable(booktabs = TRUE,
               format = "html",
               caption = "Trees",
             caption.short = "Trees"
             # col.names = c("Subject/Relevance",
             #              "Desc.",
             #              "Type",
             #              "Obsv. (n)",
             #              "Resolution (m)",
             #              "Source")
             ) %>% 
  kableExtra::row_spec(13, extra_css = "border-bottom: 1px solid")

@@

```{r fig-density, fig.cap="Empirical density distribution of all individuals within the presented genera along the UHI continuum. UHI intensities were extracted for each tree location, and the distribution hence represents the first detailed overview of the exposure of Berlin's trees to urban heat loading. The black line represents the density across all three categories. Insets represent corresponding tree totals.",fig.width = 12, fig.height = 12, out.width="100%"}
knitr::include_graphics(here::here("analysis/figures/plot_02_genus_UHI_dens.png"))

@@

The effect of UHI loading on absolute growth potential varies between genera and species (Fig.$~$\@ref(fig:fig-lmestat)).
Most notably, *Quercus*, the 3rd-most frequent genera, shows decreased absolute growth with increased UHI,
while the most frequent genera, *Tilia* clearly features enhanced growth.
Note, that the estimated effect sizes here are linear.
Temperature may excert a non-linear control on absolute growth, however, and hence, applying a  method able to capture may result in altered presented effect sizes.
Additionally, if temperatures increase in the future under climate warming current future increases in under climate warming, any non-linear effects may become more enhanced, stressing the need for a more flexible model fit and structure.

alter the (preliminary) results substantially

@@

- incorporating more pertinent covariates as dependent variables in the linear mixed model
- testing multiple model structures with formal model selection procedures
- checking model resiudals for spatial auto-correlation and accounting for it where necessary to ensure unbiased estimates of effect sizes
- repeating the above with a hierarchical GAM (i.e. GAMM) to allow for:  
  * estimating continuous prediction surfaces for UHI impacts on individual species' growth (similar to results in Figure$~$\@ref(fig:fig-lmestat)) under recent conditions
  * estimating absolute, species-specific growth potential under increased temperatures and UHI loading under climate change, ideally based on climate simulations (otherwise step-wise increases based on RCP scenarios) for the key species. Note, that  complications presented by 'out-of-sample' predictions will be addressed.
  * assess potential age-dependent UHI impacts on individual species.

# Acknowledgements

<!-- The following line inserts a page break  -->
`r pgbreak()`

>>>>

---
title: "Species-specific effects of the Urban Heat Island on tree growth across Berlin"
author:
  - Alexander G. Hurley$^{a *}$, Ingo Heinrich$^a$, ...
  - $^a$ *Geoforschungszentrum Potsdam, Section 4.3, Germany*
  - "* Corresponding author: hurley@gfz-potsdam.de"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template.docx" # Insert path for the DOCX file
      pandoc_args:
      - --lua-filter=../templates/scholarly-metadata.lua
      - --lua-filter=../templates/author-info-blocks.lua
      - --lua-filter=../templates/pagebreak.lua
    bookdown::html_document2:
      fig_caption: yes
      toc: true
    bookdown::pdf_document2:
      fig_caption: yes
      toc: yes
header-includes:
- \usepackage{booktabs}
- \usepackage{longtable}
- \usepackage{array}
- \usepackage{multirow}
- \usepackage{wrapfig}
- \usepackage{float}
- \usepackage{pdflscape}
- \usepackage{tabu}
- \usepackage{threeparttable}
- \usepackage[normalem]{ulem}
bibliography: uhi_references.bib
csl: "../templates/journal-of-archaeological-science.csl" # Insert path for the bib-style
abstract: |
  **This document outlines the rationale for an analysis of tree growth (potential) and its relationship with the Urban Heat Island (UHI) effect in Berlin using an extensive, publicly available data set. It introduces preliminary results and provides an outlook for up-coming and potential work.**
keywords: |
  Uferbaeume; Anlagenbaeume; Strassenbaeume; 
highlights: |
  We will see. 
---

@@

To complement (existing) detailed dendroecological analyses of climate-growth relationships in Berlin for key species, we propose inferring growth modulation from a large data set in excess of 650000 individuals provided by the Berlin Senate Administration (Senatsverwaltung).
This data set contains information on location, species, trunk diameter (at breast height; $DBH$; see Tab.$~$\@ref(tab:tab-tree-overview)), and height, amongst other variables for street and park trees.

```{r tab-tree-overview}
loadd(overview_table)

caption_ov <- "Available records by categoryin entire data set (n), and those with age and $DBH$ entries (n$_{full}$)."
caption_ov_short <- "Available records in berlin tree database."

if (grepl("docx", out_format) ||
    knitr::is_latex_output() ) {
  table_ov <- overview_table %>% 
  knitr::kable(booktabs = TRUE,

             caption = caption_ov,
             caption.short = caption_ov_short,
             format = "latex",
             # col.names = c("Subject/Relevance",
             #              "Desc.",
             #              "Type",
             #              "Obsv. (n)",
             #              "Resolution (m)",
             #              "Source")
             escape = FALSE)

} else {
table_ov <- overview_table %>% 
  knitr::kable(booktabs = TRUE,
               format = "html",

             caption = caption_ov,
             caption.short = caption_ov_short,
             # col.names = c("Subject/Relevance",
             #              "Desc.",
             #              "Type",
             #              "Obsv. (n)",
             #              "Resolution (m)",
             #              "Source")
             escape = FALSE) 
}

 table_ov %>% 
  kableExtra::kable_styling(full_width = TRUE, 
                            font_size = 11,
                            latex_options = c("hold_position")) %>% 
  kableExtra::row_spec(0, bold = TRUE, font_size = 12)

In a space-for-time substitution, growth of individual species can be assessed across the entire cite of Berlin, and related to effects of the UHI, while accounting for other location-specific factors, such as street characteristics, development intensity, available soil volume, etc.
Comparable applications are found, for example, in @quigley2004 and @pretzsch2017.
The former  inferred absolute growth potential for species across successional groups, and between rural and urban species, yet lacked spatially-explicit effect size estimates or predictions of maximum potential;
@pretzsch2017 applied linear hierarchical models to infer growth modulation for different cities, time periods and urban $vs.$ rural locations while accounting for stand-level variability.

**By contrast, we propose applying a statistical model that is fully spatially explicit, while also allowing to account for the nested nature of the data set (e.g. streets and districts) as well as other pertinent factors using hierarchical, generalized additive models (see Section$~$\@ref(sec:methods).
This also allows to infer the absolute growth potential of a species given, for example, a specific location, age or UHI magnitude.**
Tree-level growth data, however, is paramount in validating such relationships, and its inclusion in the model would also allow incorporating effects of varying climate over time.

@@

Currently, $DBH$ has been modelled using a hierarchical linear model (linear mixed effects model) with `lme4` [@bates2015] in @rcoreteam2020 (see Section$~$\@ref(sec:results)).
The general form of this model is:

\begin{equation}
Y_{i,j} = (\beta_0 + b_{0,i,j}) + (\beta_1 + b_{1,i,j}) \cdot x_i  + \epsilon_{i,j},
(\#eq:lme)
\end{equation}
where $\beta_0$ is the intercept with its random component $b_0$, and $\beta_1$ the slope with its random component $b_1$.
The random errors are assumed i.i.d. and distributed as $b \sim \mathcal{N}(0, \tau ^2)$. 
The model for which results are presented in Figure$~$\@ref(fig:fig-lmestat) estimates $DBH$ from tree age and the local UHI intensity as continuous covariates with random slopes and intercepts for each species;
note, that for computational efficiency each genera was modelled seperately.
Further, models were established for the three most abundant species per genera with at least 1000 individuals.

We caution, that 1) the presented model results are preliminary, 2) initial diagnostics indicate that modification of the current model structure (e.g. by accounting for individual-level variability) may be required and 3) that further, time-intensive data cleaning is necessary. 
First tests, however, showed that while the magnitude of effects changes with different model structures, their direction appears stable.

@@

Additional remote sensing data for local context, higher resolution UHI data, and most important, temporally-resolved tree growth data are still required and/or would greatly improve the impact and confidence in effect size estimates.
See Tab.$~$\@ref(tab:tab-data-req) for details.

@@

caption_data_req <- "Data requirements for analysis, including currently available/accessed and required/desired data."
caption_data_req_short <- "Data requirements for analysis."

@@

             caption = caption_data_req,
             caption.short = caption_data_req_short,
             format = "latex",
             col.names = c("Subject / Relevance",
                          "Desc.",
                          "Type",
                          "Obsv. (n)",
                          "Resolution (m)",
                          "Source")) %>% 
  kableExtra::kable_styling(full_width = TRUE, 
                            font_size = 11) %>% 
  kableExtra::row_spec(0, bold = TRUE, font_size = 12) %>% 
  kableExtra::row_spec(9,italic =  TRUE) %>% 
  kableExtra::pack_rows("available or accessed", 1, 4) %>%
  kableExtra::pack_rows("required and/or desired", 5, 9) %>% kableExtra::landscape()

@@

# Preliminary results {#sec:results}

Tree locations are clustered and structured based on their category, i.e. riparian, street and park trees (Fig.$~$\@ref(fig:fig-tree-overview-map)).
Planting in space and time shows species-specific patterns (by districts), often related to major events, such as the start and end of armed and/or political conflict.
Table$~$\@ref(tab:tab-age) shows the binned distribution of genera across age classes.

@@

age_tables <- age_tables %>% 
  dplyr::arrange(desc(n_total))
colnames(age_tables)[c(1,7,8, 9)] <- c("Genera", "150+", "Total (n)", "Missing (n)")

@@

caption_trees <- "Binned age-distribution for street and park trees. Missing values feature no"
caption_trees_short <- "Binned age-distribution for street and park trees."

if (grepl("docx", out_format) ||
    knitr::is_latex_output() ) {
  table_dat_req_kable <- age_tables %>% 
  knitr::kable(booktabs = TRUE,
               format = "latex",
             caption = caption_trees,
             caption.short = caption_trees_short
             # col.names = c("Subject/Relevance",
             #              "Desc.",
             #              "Type",
             #              "Obsv. (n)",
             #              "Resolution (m)",
             #              "Source")
             ) %>% 
  kableExtra::row_spec(13, hline_after = TRUE)

} else {
table_dat_req_kable <- age_tables %>% 
  knitr::kable(booktabs = TRUE,
               format = "html",
              caption = caption_trees,
             caption.short = caption_trees_short
             # col.names = c("Subject/Relevance",
             #              "Desc.",
             #              "Type",
             #              "Obsv. (n)",
             #              "Resolution (m)",
             #              "Source")
             ) %>% 
  kableExtra::row_spec(13, extra_css = "border-bottom: 1px solid")

@@

```{r fig-density, fig.cap="Empirical density distribution of all individuals within the presented genera along the UHI continuum. UHI intensities were extracted for each tree location, and the distribution hence represents the first detailed overview of the exposure of Berlin's trees to urban heat loading. The black line is the density across all three categories. Insets show corresponding tree totals.",fig.width = 12, fig.height = 12, out.width="100%"}
knitr::include_graphics(here::here("analysis/figures/plot_02_genus_UHI_dens.png"))

@@

The effect of UHI loading on absolute growth potential varies between genera and species (Fig.$~$\@ref(fig:fig-lmestat)); **note, that these results are preliminary and should be considered as template for future outputs, rather used for any inference**.
Most notably, *Quercus*, the 3rd-most frequent genera, shows decreased absolute growth with increased UHI,
while the most frequent genera, *Tilia* features contrasting relationships with UHI at species level.
The estimated effect sizes presented here are linear.
However, temperature may excert a non-linear control on absolute growth and, hence, applying a  method able to capture may result in altered presented effect sizes.
Additionally, if temperatures increase in the future under climate warming current future increases in under climate warming, any non-linear effects may become more enhanced, stressing the need for a more flexible model fit and structure.

@@

- validating the database with independent observations
- incorporating more pertinent covariates as dependent variables in the linear mixed model, especially incremental tree growth
- testing multiple model structures with formal model selection procedures
- checking model resiudals for spatial auto-correlation and accounting for it where necessary to ensure unbiased estimates of effect sizes
- repeating the above with a hierarchical GAM (i.e. GAMM) to allow for:  
  * estimating continuous prediction surfaces for UHI impacts on individual species' growth (similar to results in Figure$~$\@ref(fig:fig-lmestat)) under recent conditions
  * estimating absolute, species-specific growth potential under increased temperatures and UHI loading under climate change, ideally based on climate simulations (otherwise step-wise increases based on RCP scenarios) for the key species. Note, that  complications presented by 'out-of-sample' predictions will be addressed.
  * assess potential age-dependent UHI impacts on individual species.

<!-- # Acknowledgements -->

<!-- <!-- The following line inserts a page break  --> 
<!-- `r pgbreak()` -->
